---
fontsize: 12pt
header-includes:
- \usepackage{amsmath}
- \usepackage{mathtools}
- \usepackage{verbatim}
- \usepackage{setspace}
- \usepackage{lineno}
- \linenumbers
- \usepackage{float} \floatplacement{figure}{H}
- \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\thesection.\arabic{table}}
  \setcounter{figure}{0} \renewcommand{\thefigure}{S\thesection.\arabic{figure}}}
bibliography: bwrisk.bib
output:
  bookdown::pdf_document2:
    toc: no
    pandoc_args: [ "--csl", "marine-mammal-science.csl"]
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyr)
library(dplyr)
library(knitr)

load("../Data/rlDataN.RData")
rlData <- rlDataN

baseline <- filter(rlData, Period=="00") 
ships <- filter(rlData, Period=="02")
sonar <- filter(rlData, Period == "12")
```

\begin{center}

\Large{
\textbf{
\uppercase{Quantifying the response of Blainville's beaked whales to US Naval sonar exercises in Hawaii}}}


\normalsize
\hfill\break  
\textbf{Eiren K. Jacobson\textsuperscript{1}, E. Elizabeth Henderson\textsuperscript{2}, David L. Miller\textsuperscript{1}, Cornelia S. Oedekoven\textsuperscript{1}, David J. Moretti\textsuperscript{3}, Len Thomas\textsuperscript{1}}
  
\small
\textsuperscript{1} \textit{Centre for Research into Ecological and Environmental Modelling, School of Mathematics and Statistics, University of St Andrews, St Andrews, Scotland}  

\textsuperscript{2} \textit{Naval Information Warfare Center Pacific, San Diego, CA, USA}  

\textsuperscript{3} \textit{Naval Undersea Warfare Center, Newport, RI, USA}  

\normalsize

\textbf{Correspondence:}  

Eiren Jacobson \\
University of St Andrews \\ 
The Observatory  \\
Buchanan Gardens  \\
St Andrews \\
Fife \\
KY16 9LZ \\  
Scotland \\

Email: eiren.jacobson@st-andrews.ac.uk

\normalsize
\hfill\break   
\textbf{Draft `r format(Sys.time(), "%d %B %Y")`}
\end{center}

\newpage
\doublespace

## Abstract {.unlisted .unnumbered}

Behavioral responses of beaked whales (family Ziphiidae) to naval use of mid-frequency active sonar (MFAS) have been quantified for some species and regions. We describe the effects of MFAS on the probability of detecting diving groups of Blainville's beaked whales on the US Navy Pacific Missile Range Facility (PMRF) in Hawaii and compare our results to previously published results for the same species at the Atlantic Undersea Test and Evaluation Center (AUTEC) in the Bahamas.  We used passive acoustic data collected at bottom-mounted hydrophones before and during six naval training exercises at PMRF along with modelled sonar received levels to describe the effect of training and MFAS on foraging groups of Blainville's beaked whales.  We used a multi-stage generalized additive modelling approach to control for the underlying spatial distribution of vocalizations under baseline conditions.  At a MFAS received level of 150\ dB\ re\ 1\ $\mu$Pa the probability of detecting groups of Blainville's beaked whales decreased by 78\% (95\% CI 62\%-100\%) compared to periods when general training activity was ongoing and by 92\% (95\% CI 87\%-100\%) compared to baseline conditions.  Our results indicate a more pronounced response to naval training and MFAS than has been previously reported.  [196/200]

### KEYWORDS {.unlisted .unnumbered}

Blainville's beaked whales, \textit{Mesopolodon densirostris}, mid-frequency active sonar, passive acoustic data, behavioral response, generalized additive model 



\newpage

```{r child = '01_Intro.Rmd'}
```

```{r child = '02_Methods.Rmd'}
```

```{r child = '03_Results.Rmd'}
```

```{r child = '04_Discussion.Rmd'}
```

### Acknowledgements {.unlisted .unnumbered}

This study was funded by the US Navy Living Marine Resources Program  (Contract No. N39430-17-P-1983).

### Authors' Contributions {.unlisted .unnumbered}

Conceptualization: E.E.H., D.J.M, L.T.  
Data curation: E.K.J., E.E.H.  
Formal analysis: E.K.J., E.E.H., C.S.O.  
Funding acquisition: E.E.H., D.J.M., L.T.  
Investigation: E.E.H.  
Methodology: E.K.J., E.E.H., D.L.M., C.S.O., L.T.  
Software: E.K.J., D.L.M., C.S.O.  
Supervision: L.T.  
Visualization: E.K.J.  
Writing -- original draft: E.K.J., E.E.H., D.L.M.  
Writing -- review & editing: E.K.J, E.E.H., D.L.M., C.S.O., D.J.M., L.T.  

### ORCID {.unlisted .unnumbered}

Eiren K. Jacobson: https://orcid.org/0000-0003-0147-8367  
David L. Miller: https://orcid.org/0000-0002-9640-6755  
Cornelia S. Oedekoven: https://orcid.org/0000-0002-5610-7814  
Len Thomas: https://orcid.org/0000-0002-7436-067X  

\newpage

## References {.unlisted .unnumbered}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent

<div id="refs"></div>

\newpage

\beginsupplement
\setcounter{section}{0}

\newpage
\refstepcounter{section}


```{r child = '05_S1_Uncertainty.Rmd'}
```

\newpage
\refstepcounter{section}
\setcounter{figure}{0}

```{r child = '07_S3_FigsTabs.Rmd'}
```

\newpage
\refstepcounter{section}
\setcounter{figure}{0}

```{r child = '08_S4_SingleGAM.Rmd'}
```

\newpage